% Garrett Benoit
% Thesis class, based on report
\documentclass
[
	%short,	% omit front matter
	%draft, 	% draft heading, don't load graphics, etc.
]
{thesis}

% Optional additional packages
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{chicago}
\usepackage{listings}
\usepackage{color}
\usepackage{verbatim}
\usepackage{subcaption}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
	language=Java,
	aboveskip=3mm,
	belowskip=3mm,
	showstringspaces=false,
	columns=flexible,
	basicstyle={\small\ttfamily},
	numbers=none,
	numberstyle=\tiny\color{gray},
	keywordstyle=\color{blue},
	commentstyle=\color{dkgreen},
	stringstyle=\color{mauve},
	breaklines=true,
	breakatwhitespace=true,
	tabsize=3
}

% Thesis parameters
\title{Alternatives using the Leap Motion to extend Mid-Air Word-Gesture Keyboards}
\author{Garrett Benoit}
\holding{B.S.}
\seeking{M.S.}
\degree{Master of Science}
\date{October 2015}

% Administration
\graduateDean{J. Larry Lyon, Ph.D.}
\department{Department of Computer Science}
\departmentChair{Gregory D. S............................................peegle, Ph.D.}

% Supervisor (adviser / mentor)
\supervisor{Michael G. Poor, Ph.D.}
\supervisorTitle{Faculty Adviser}

%  Readers (committee members)
\readerOne{Bill Poucher, Ph.D.}
\readerTwo{Brian Garner, Ph.D.}
%\readerThree{...}
%\readerFour{...}
%\readerFive{...}

% Abstract
\abstract{Lately, the application of touchless, mid-air, gesture-based interactions has risen in popularity for a variety of fields (e.g., augmented and virtual reality, gaming, etc.). With this newfound, wide-spread application comes the need for effecient, mid-air, gesture-based text-entry; the solution: apply word-gesture keyboards to mid-air text-entry. Markussen et al. used this technique for the first time with the inception of Vulture \cite{ref_vulture}, the first mid-air, word-gesture keyboard, providing a mid-air text-entry rate of 28.1 Words Per Minute, the fastest rate yet. This study builds on the findings of Markussen et al. and presents alternative means of word separation for mid-air, word-gesture keyboards; as a result, exploring and identifying the problems of new techniques and presenting the possible solutions. Results revealed that a bimodal approach holds great promise, reaching a mean text-entry rate of 15.8 Words Per Minutes for a single session with no training.}

% Acknowledgments (optional)
\acknowledgements{To Dr. G. Michael Poor for allowing me to explore this idea in depth, guiding me to a path of success, and for putting up with me the whole time. Alvin Jude Hari Haran and Darren Guinness for their invaluable input that helped shaped the direction of this study. To all of those in the HCI Lab who helped with my pilot studies! To all the volunteers who made this research possible by generously donating their time. Finally, to my sister, Heather Benoit, and other family members who inspired me to go to Baylor and seek a greater future.}

% Dedication (optional)
\dedication{\it To those who pioneer the future.}

% List of figures (optional)
\makeListOfFigures

% List of tables (optional)
\makeListOfTables

% Document body
\begin{document}

	% Chapters
	\include{ch1}
	\include{ch2}
	\include{ch3}
	\include{ch4}
	\include{ch5}
	\include{ch6}
	
	% Appendices (optional)
	% TODO: make the thesisAppendixPage automagic if the \begin{appendices} exists
	\thesisAppendixPage 
	\begin{appendices}
		\include{appendixA}
		\include{appendixB}
	\end{appendices}
	
	% Bibliography
	\bibliographystyle{chicago} % or whatever style your department uses
	\bibliography{thesis}
	
\end{document}
